Coinfluence: The Role of Currency Numerosity in Consumer Decision Making

### 2024-04-17 paper.qmd
done

---
title: "Effect of the Nominal Value of Foreign Currencies in valuing a good"
date: 2024-04-16
author:
  - name: Konrad Kopp
    email: konrad.kopp@stcatz.ox.ac.uk
    affiliation: 
      - name: University of Oxford
abstract: > 
  Research into the effect of using foreign currencies on consumption behaviour suggests that consumers systematically under- or overspend based on the nominal (face) value of the foreign currency. However, previous research is undecided on which direction this effect has, with the two most prominent models leading to different and mutually exclusive outcomes. This paper considers the different models and tests their predictive and explanatory accuracy using a study. It finds that individuals tend to underspend when the foreign currency is less numerous than (a fraction of) the home currency and overspend when the opposite is the case. The paper discusses which model is most accurate at predicting and explaining this effect and considers limitations as well as opportunities for future research.
license: "CC0"
copyright: 
  holder: Konrad Kopp
  year: 2024
keywords:
  - Behavioural Economics
  - Face Value Effect
  - Consumer Choice Theory
---

@@

Experience suggests that we treat foreign currencies differently to our home currency and have difficulty adjusting to them. Previous research, such as the "money illusion" effect [@Shafir1997] suggests that individuals are biased towards the nominal (or face value) of prices. This bias has also been explored in the context foreign currencies, with research finding systematic effects that arise based on the numerousity of values in a foreign currency compared to individuals' home currencies. However, previous research is split on how and why this effect occurs.

@@

First, I present the theoretical frameworks that have been outlined by previous research in a standard form that makes it easy to compare and contrast them. These frameworks are the standard model described in consumer choice theory, the model suggested by Raghubir and Srivastava and the model suggested by Wertenbroch, Soman and Chattopadhyay. Then, I consider what each of these frameworks would predict in a concrete case and show that all three models come to different and mutually exclusive conclusions. In order to test which of these frameworks is most accurate in predicting observed behaviour, I conduct a study in which two groups each formulate their WTP for a specific good in an incentive compatible way, with the only difference between the groups being the exchange rate of the currency used to their home currency and, as a result, the numerousity of the values of their budget and bid. Analysing the study, I find that consumers tend to underspend when the foreign currency is a fraction and overspend when the foreign currency is a multiple, which is predicted by the model proposed by Wertenbroch, Soman and Chattopadhyay [@Wertenbroch2007]. However, due to the limited sample size of the study, this result is not statistically significant and does not have significant statistical power. In the discussion, I evaluate the implications of the findings for the theoretical frameworks outlined, consider alternative explanations for the perceived effect and argue that the model proposed by Wertenbroch, Soman and Chattopadhyay is the most accurate in predicting and describing consumer behaviour when using foreign currencies. Finally, I will consider the limitations of this paper and opportunities for further research.

@@

Since the aim of the study was to determine whether there were any systematic differences in WTP between consumers when the only differing factor is the numerousity of the currency used, I decided to not include a control group that bid on the item using Pounds, which was the design of some of the experiments run by Raghubir and Srivastava [@Raghubir2002]. This allowed for a simpler design and more power given the same sample size. Further, unlike the studies run by Raghubir and Srivastava [@Raghubir2002] and Wertenbroch, Soman and Chattopadhyay [@Wertenbroch2007], I did not use a fiat currency or mix of fiat and virtual currencies, but used only one virtual currency for both groups. On the one hand, this simplification was due to a focus of wether the effects described in previous papers also held in similar ways for virtual currencies. On the other hand, it also allowed me to preclude any biases or previous experience that participants might have had towards or with certain currencies. Raghubir and Srivastava [@Raghubir2002] acknowledge that "the face value effect is due to the accessibility and perceptual salience of the face value of the foreign currency ... [which] is likely to depend on the extent to which an individual has the opportunity or the time available to process exchange rate information and/or has experience in using a particular foreign currency". Further, Alter and Oppenheimer [@Alter2008] show that people use "familiarity and fluency" when valuing goods in different currencies. To bias participants the least, I chose the relatively nondescript name for a fictional, virtual currency: BlueCoin.

@@

The study conducted found that the fraction group bid less on average than the multiple group. The difference in the means across the two groups is around £6, or over 10% of the budget, signifying that this difference occurs due to a systematic difference in how each group calculated their value in the foreign currency BlueCoin rather than random chance. The only difference between the groups was the exchange rate of BlueCoin to the Pound and, as a result, the numerousity of the nominal values of the budget and the bids. This would suggest that this difference in bids occurs based on whether the foreign currency is a fraction or multiple of the home currency.

@@

Apart from the limitations, the major theoretical implication of this paper is the finding that the model proposed by Wertenbroch, Soman and Chattopadhyay [@Wertenbroch2007] seems to be most accurate in predicting and explaining consumer behaviour when eliciting a WTP on a good in a foreign currncy. In line with these authors, this conclusion contradicts the findings of the "face value effect" by Raghubir and Srivastava [@Raghubir2002] and subsequently by Fang [@Fang2019]. Further, this conclusion also implies that, contrary to traditional consumption theory, individuals formulate their WTP in relation to the difference between their budget and the valuation of a good rather than its' ratio.

>>>>

---
title: "Coinfluence: The Role of Currency Numerosity in Consumer Decision Making"
date: 2024-04-16
author:
  - name: Konrad Kopp
    affiliation: 
      - name: University of Oxford
abstract: > 
  Research into the effect of using foreign currencies on consumption behaviour suggests that consumers systematically under- or overspend based on the nominal (face) value of the foreign currency. However, previous research is undecided on which direction this effect has, with the two most prominent models leading to different and mutually exclusive outcomes. This paper considers the different models and tests their predictive and explanatory accuracy using a study. It finds that individuals tend to underspend when the foreign currency is less numerous than (a fraction of) the home currency and overspend when the opposite is the case. The paper discusses which model is most accurate at predicting and explaining this effect and considers limitations as well as opportunities for future research.
license: "CC0"
keywords:
  - Behavioural Economics
  - Face Value Effect
  - Consumer Choice Theory
---

@@

Experience suggests that we treat foreign currencies differently to our home currency and have difficulty adjusting to them. Previous research, such as the "money illusion" effect [@Shafir1997] suggests that individuals are biased towards the nominal (or face value) of prices. This bias has also been explored in the context foreign currencies, with research finding systematic effects that arise based on the numerosity of values in a foreign currency compared to individuals' home currencies. However, previous research is split on how and why this effect occurs.

@@

First, I present the theoretical frameworks that have been outlined by previous research in a standard form that makes it easy to compare and contrast them. These frameworks are the standard model described in consumer choice theory, the model suggested by Raghubir and Srivastava and the model suggested by Wertenbroch, Soman and Chattopadhyay. Then, I consider what each of these frameworks would predict in a concrete case and show that all three models come to different and mutually exclusive conclusions. In order to test which of these frameworks is most accurate in predicting observed behaviour, I conduct a study in which two groups each formulate their WTP for a specific good in an incentive compatible way, with the only difference between the groups being the exchange rate of the currency used to their home currency and, as a result, the numerosity of the values of their budget and bid. Analysing the study, I find that consumers tend to underspend when the foreign currency is a fraction and overspend when the foreign currency is a multiple, which is predicted by the model proposed by Wertenbroch, Soman and Chattopadhyay [@Wertenbroch2007]. However, due to the limited sample size of the study, this result is not statistically significant and does not have significant statistical power. In the discussion, I evaluate the implications of the findings for the theoretical frameworks outlined, consider alternative explanations for the perceived effect and argue that the model proposed by Wertenbroch, Soman and Chattopadhyay is the most accurate in predicting and describing consumer behaviour when using foreign currencies. Finally, I will consider the limitations of this paper and opportunities for further research.

@@

Since the aim of the study was to determine whether there were any systematic differences in WTP between consumers when the only differing factor is the numerosity of the currency used, I decided to not include a control group that bid on the item using Pounds, which was the design of some of the experiments run by Raghubir and Srivastava [@Raghubir2002]. This allowed for a simpler design and more power given the same sample size. Further, unlike the studies run by Raghubir and Srivastava [@Raghubir2002] and Wertenbroch, Soman and Chattopadhyay [@Wertenbroch2007], I did not use a fiat currency or mix of fiat and virtual currencies, but used only one virtual currency for both groups. On the one hand, this simplification was due to a focus of wether the effects described in previous papers also held in similar ways for virtual currencies. On the other hand, it also allowed me to preclude any biases or previous experience that participants might have had towards or with certain currencies. Raghubir and Srivastava [@Raghubir2002] acknowledge that "the face value effect is due to the accessibility and perceptual salience of the face value of the foreign currency ... [which] is likely to depend on the extent to which an individual has the opportunity or the time available to process exchange rate information and/or has experience in using a particular foreign currency". Further, Alter and Oppenheimer [@Alter2008] show that people use "familiarity and fluency" when valuing goods in different currencies. To bias participants the least, I chose the relatively nondescript name for a fictional, virtual currency: BlueCoin.

@@

The study conducted found that the fraction group bid less on average than the multiple group. The difference in the means across the two groups is around £6, or over 10% of the budget, signifying that this difference occurs due to a systematic difference in how each group calculated their value in the foreign currency BlueCoin rather than random chance. The only difference between the groups was the exchange rate of BlueCoin to the Pound and, as a result, the numerosity of the nominal values of the budget and the bids. This would suggest that this difference in bids occurs based on whether the foreign currency is a fraction or multiple of the home currency.

@@

Apart from the limitations, the major theoretical implication of this paper is the finding that the model proposed by Wertenbroch, Soman and Chattopadhyay [@Wertenbroch2007] seems to be most accurate in predicting and explaining consumer behaviour when eliciting a WTP on a good in a foreign currency. In line with these authors, this conclusion contradicts the findings of the "face value effect" by Raghubir and Srivastava [@Raghubir2002] and subsequently by Fang [@Fang2019]. Further, this conclusion also implies that, contrary to traditional consumption theory, individuals formulate their WTP in relation to the difference between their budget and the valuation of a good rather than its' ratio.

### 2024-04-17 paper.tex
done

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Effect of the Nominal Value of Foreign Currencies in valuing a good},
  pdfauthor={Konrad Kopp},
  pdflang={en-GB},
  pdfkeywords={Behavioural Economics, Face Value Effect, Consumer Choice
Theory},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Effect of the Nominal Value of Foreign Currencies in valuing a
good}

  \author{Konrad Kopp}
            \affil{%
                  University of Oxford
              }

@@

Experience suggests that we treat foreign currencies differently to our
home currency and have difficulty adjusting to them. Previous research,
such as the ``money illusion'' effect (Shafir, Diamond, and Tversky
1997) suggests that individuals are biased towards the nominal (or face
value) of prices. This bias has also been explored in the context
foreign currencies, with research finding systematic effects that arise
based on the numerousity of values in a foreign currency compared to
individuals' home currencies. However, previous research is split on how
and why this effect occurs.

@@

First, I present the theoretical frameworks that have been outlined by
previous research in a standard form that makes it easy to compare and
contrast them. These frameworks are the standard model described in
consumer choice theory, the model suggested by Raghubir and Srivastava
and the model suggested by Wertenbroch, Soman and Chattopadhyay. Then, I
consider what each of these frameworks would predict in a concrete case
and show that all three models come to different and mutually exclusive
conclusions. In order to test which of these frameworks is most accurate
in predicting observed behaviour, I conduct a study in which two groups
each formulate their WTP for a specific good in an incentive compatible
way, with the only difference between the groups being the exchange rate
of the currency used to their home currency and, as a result, the
numerousity of the values of their budget and bid. Analysing the study,
I find that consumers tend to underspend when the foreign currency is a
fraction and overspend when the foreign currency is a multiple, which is
predicted by the model proposed by Wertenbroch, Soman and Chattopadhyay
(Wertenbroch, Soman, and Chattopadhyay 2007). However, due to the
limited sample size of the study, this result is not statistically
significant and does not have significant statistical power. In the
discussion, I evaluate the implications of the findings for the
theoretical frameworks outlined, consider alternative explanations for
the perceived effect and argue that the model proposed by Wertenbroch,
Soman and Chattopadhyay is the most accurate in predicting and
describing consumer behaviour when using foreign currencies. Finally, I
will consider the limitations of this paper and opportunities for
further research.

@@

Since the aim of the study was to determine whether there were any
systematic differences in WTP between consumers when the only differing
factor is the numerousity of the currency used, I decided to not include
a control group that bid on the item using Pounds, which was the design
of some of the experiments run by Raghubir and Srivastava (Raghubir and
Srivastava 2002). This allowed for a simpler design and more power given
the same sample size. Further, unlike the studies run by Raghubir and
Srivastava (Raghubir and Srivastava 2002) and Wertenbroch, Soman and
Chattopadhyay (Wertenbroch, Soman, and Chattopadhyay 2007), I did not
use a fiat currency or mix of fiat and virtual currencies, but used only
one virtual currency for both groups. On the one hand, this
simplification was due to a focus of wether the effects described in
previous papers also held in similar ways for virtual currencies. On the
other hand, it also allowed me to preclude any biases or previous
experience that participants might have had towards or with certain
currencies. Raghubir and Srivastava (Raghubir and Srivastava 2002)
acknowledge that ``the face value effect is due to the accessibility and
perceptual salience of the face value of the foreign currency \ldots{}
{[}which{]} is likely to depend on the extent to which an individual has
the opportunity or the time available to process exchange rate
information and/or has experience in using a particular foreign
currency''. Further, Alter and Oppenheimer (Alter and Oppenheimer 2008)
show that people use ``familiarity and fluency'' when valuing goods in
different currencies. To bias participants the least, I chose the
relatively nondescript name for a fictional, virtual currency: BlueCoin.

@@

The study conducted found that the fraction group bid less on average
than the multiple group. The difference in the means across the two
groups is around £6, or over 10\% of the budget, signifying that this
difference occurs due to a systematic difference in how each group
calculated their value in the foreign currency BlueCoin rather than
random chance. The only difference between the groups was the exchange
rate of BlueCoin to the Pound and, as a result, the numerousity of the
nominal values of the budget and the bids. This would suggest that this
difference in bids occurs based on whether the foreign currency is a
fraction or multiple of the home currency.

@@

Apart from the limitations, the major theoretical implication of this
paper is the finding that the model proposed by Wertenbroch, Soman and
Chattopadhyay (Wertenbroch, Soman, and Chattopadhyay 2007) seems to be
most accurate in predicting and explaining consumer behaviour when
eliciting a WTP on a good in a foreign currncy. In line with these
authors, this conclusion contradicts the findings of the ``face value
effect'' by Raghubir and Srivastava (Raghubir and Srivastava 2002) and
subsequently by Fang (Fang 2019). Further, this conclusion also implies
that, contrary to traditional consumption theory, individuals formulate
their WTP in relation to the difference between their budget and the
valuation of a good rather than its' ratio.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Coinfluence: The Role of Currency Numerosity in Consumer Decision Making},
  pdflang={en-GB},
  pdfkeywords={Behavioural Economics, Face Value Effect, Consumer Choice
Theory},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Coinfluence: The Role of Currency Numerosity in Consumer Decision
Making}

  \author{1054398}
            \affil{%
                  University of Oxford
              }

@@

Experience suggests that we treat foreign currencies differently to our
home currency and have difficulty adjusting to them. Previous research,
such as the ``money illusion'' effect (Shafir, Diamond, and Tversky
1997) suggests that individuals are biased towards the nominal (or face
value) of prices. This bias has also been explored in the context
foreign currencies, with research finding systematic effects that arise
based on the numerosity of values in a foreign currency compared to
individuals' home currencies. However, previous research is split on how
and why this effect occurs.

@@

First, I present the theoretical frameworks that have been outlined by
previous research in a standard form that makes it easy to compare and
contrast them. These frameworks are the standard model described in
consumer choice theory, the model suggested by Raghubir and Srivastava
and the model suggested by Wertenbroch, Soman and Chattopadhyay. Then, I
consider what each of these frameworks would predict in a concrete case
and show that all three models come to different and mutually exclusive
conclusions. In order to test which of these frameworks is most accurate
in predicting observed behaviour, I conduct a study in which two groups
each formulate their WTP for a specific good in an incentive compatible
way, with the only difference between the groups being the exchange rate
of the currency used to their home currency and, as a result, the
numerosity of the values of their budget and bid. Analysing the study, I
find that consumers tend to underspend when the foreign currency is a
fraction and overspend when the foreign currency is a multiple, which is
predicted by the model proposed by Wertenbroch, Soman and Chattopadhyay
(Wertenbroch, Soman, and Chattopadhyay 2007). However, due to the
limited sample size of the study, this result is not statistically
significant and does not have significant statistical power. In the
discussion, I evaluate the implications of the findings for the
theoretical frameworks outlined, consider alternative explanations for
the perceived effect and argue that the model proposed by Wertenbroch,
Soman and Chattopadhyay is the most accurate in predicting and
describing consumer behaviour when using foreign currencies. Finally, I
will consider the limitations of this paper and opportunities for
further research.

@@

Since the aim of the study was to determine whether there were any
systematic differences in WTP between consumers when the only differing
factor is the numerosity of the currency used, I decided to not include
a control group that bid on the item using Pounds, which was the design
of some of the experiments run by Raghubir and Srivastava (Raghubir and
Srivastava 2002). This allowed for a simpler design and more power given
the same sample size. Further, unlike the studies run by Raghubir and
Srivastava (Raghubir and Srivastava 2002) and Wertenbroch, Soman and
Chattopadhyay (Wertenbroch, Soman, and Chattopadhyay 2007), I did not
use a fiat currency or mix of fiat and virtual currencies, but used only
one virtual currency for both groups. On the one hand, this
simplification was due to a focus of wether the effects described in
previous papers also held in similar ways for virtual currencies. On the
other hand, it also allowed me to preclude any biases or previous
experience that participants might have had towards or with certain
currencies. Raghubir and Srivastava (Raghubir and Srivastava 2002)
acknowledge that ``the face value effect is due to the accessibility and
perceptual salience of the face value of the foreign currency \ldots{}
{[}which{]} is likely to depend on the extent to which an individual has
the opportunity or the time available to process exchange rate
information and/or has experience in using a particular foreign
currency''. Further, Alter and Oppenheimer (Alter and Oppenheimer 2008)
show that people use ``familiarity and fluency'' when valuing goods in
different currencies. To bias participants the least, I chose the
relatively nondescript name for a fictional, virtual currency: BlueCoin.

@@

The study conducted found that the fraction group bid less on average
than the multiple group. The difference in the means across the two
groups is around £6, or over 10\% of the budget, signifying that this
difference occurs due to a systematic difference in how each group
calculated their value in the foreign currency BlueCoin rather than
random chance. The only difference between the groups was the exchange
rate of BlueCoin to the Pound and, as a result, the numerosity of the
nominal values of the budget and the bids. This would suggest that this
difference in bids occurs based on whether the foreign currency is a
fraction or multiple of the home currency.

@@

Apart from the limitations, the major theoretical implication of this
paper is the finding that the model proposed by Wertenbroch, Soman and
Chattopadhyay (Wertenbroch, Soman, and Chattopadhyay 2007) seems to be
most accurate in predicting and explaining consumer behaviour when
eliciting a WTP on a good in a foreign currency. In line with these
authors, this conclusion contradicts the findings of the ``face value
effect'' by Raghubir and Srivastava (Raghubir and Srivastava 2002) and
subsequently by Fang (Fang 2019). Further, this conclusion also implies
that, contrary to traditional consumption theory, individuals formulate
their WTP in relation to the difference between their budget and the
valuation of a good rather than its' ratio.
